\chapter{Introduction}
We need program analysis techniques that can find security bugs in binary programs.
Binary program analysis is essential because code is often only available in binary form.
Even when source code is available, some security properties are more appropriately checked at the binary level.
As an example, use-after-free bugs present in source may not be present at the binary level if the compiler chooses to elide a read when a register caches its value.

Authors designed existing tools~\cite{ida, bap, bitblaze, bindead} as a sequence of analyses to perform, with the results of each previous analysis being available to the next.
In this model, an analysis author must ensure the framework calculates any information needed before their analysis, but after any analyses that could benefit their dependency's accuracy.
Representing individual analyses as nodes, with a directed edge between $n_1$ and $n_2$ indicating that $n_1$ must execute before $n_2$, this graph forms a sequence, or line.

Even in compilation, explicit staging can be cumbersome, so compilers have moved from this sequence or \emph{line} model to a dependency-based \emph{DAG} model.
The sequence model is essentially an ad-hoc form of LLVM\cite{llvm}'s pass model.
In LLVM's pass model, each pass declares what analyses it depends upon, allowing the LLVM pass manager to dynamically schedule the analyses and allow a configurable pipeline.
Considering the graph representation again, LLVM's model is a DAG.

This DAG model works well over source code, but falls short when applied to compiled code.
Jakstab~\cite{jakstab} argues for the co-dependency of value analysis and lifting. 
It takes the approach of running these two analyses to a fixed point, and provides a hook for user-defined analyses to run on each iteration.
As Jakstab does not explicitly encode dependencies between analysis, the graph of execution dependencies looks like a single cycle.
The Jakstab approach is undoubtably a step forwards.
However, it still causes unnecessary recalculation and places a form of monotonicity requirements on analyses in order to remain correct.
A natural extension then is to allow the expression of dependencies, as in LLVM, but also cycles, as in Jakstab.
We investigate this full graph model of analysis management in this thesis.

We often want co-dependent analysis to detect bugs.
Jakstab~\cite{jakstab} is one example where simply integrating two analyses (value analysis and control flow recovery/disassembly) lead to better results than IDA\cite{ida} for jump resolution.
Specifying these analyses as logical rules within the same environment would provide the same power Jakstab found by integrating them for free.
Notably, IDA misses edges which Jakstab finds.
The result is that an evaluation of a security property done over the CFG presented by IDA would miss some true positives along the undetected control flow edges.
Doop\cite{doop1} found that integrating exception analysis and points-to analysis gave them candidate target sets which were half the size on average for each object.
Extra precision of this sort translates into reduced false positives when evaluating security conditions over the program.

Our main insight is that we need an analysis framework that allows for arbitrary cycles in the dependency graph.
Further, we want a framework that is extensible so that adding new analyses is easier.
Using a logic language is a natural approach to representing such a dependency graph.

% Past work sets the stage for / encourages this
Previous work in program analysis suggests that logic languages can help to structure this problem.
% * Datalog/prolog as a program analysis tool
Datalog has previously been successfully used to analyze programs~\cite{lam2005,brumley2006,alpuente2011,doop1,bddbddb}.
Existing work has modeled a wide variety of properties from aliasing in compiled code~\cite{brumley2006} to security properties such as SQL injectability and cross site scriptability (XSS)~\cite{lam2005} as facts in a deductive database.
This work suggests a Datalog format as a potential common representation.
Dataflow analyses are also representable~\cite{mcallester2002} in this way.

% External code
Using a common extension of logic language, we can even call out to code not written as logical rules\footnote{We use a system similar to external predicates \S\ref{sec:extpred}}.
This makes it possible to repurpose previously written analyses, or to write new analyses which may not be best represented as rules.
There are still some restrictions on how such code can operate (for example, no preserved state across calls) but taking this approach gives the flexibility to be an integration system.

\begin{inset}
{\bf Thesis statement.}
A Datalog derivative with extensions for negation and external callbacks can drive multiple potentially co-dependent analyses to reason about binary programs effectively.
\end{inset}

% Road map to defending the thesis statement
% * Show the power by
%   * Implementing the integration of real analyses using the language
%   * Attempting to do so without using circumscription or aggregation, and seeing what happens.
\noindent\textbf{Co-dependent Analyses.}
We examined the co-dependent analyses of control flow recovery (\S~\ref{sec:cfg}), value analysis (\S~\ref{sec:valanal}), and alias analysis (\S~\ref{bkg:sec:alias}, \S~\ref{chap:alias}).
Datalog fits well as a structuring tool for these analyses due to its incrementality and ability to handle mutual recursion.

\noindent\textbf{Negation.}
In previous work, the framework runs an analysis ``to completion'' before proceeding.
When a later analysis uses the results of the analysis, it can use them in two ways.
The first way is direct consumption of the facts provided by the analysis.
In the case of a control flow recovery analysis, this would be treating a jump as possible because it is present in the returned CFG.
The second way is a consumption of information on what was \emph{not} returned by the analysis.
Following the control flow example, the later analysis may assume that because the control flow analysis has completed and will not be extended, any edge not present in the recovered control flow graph cannot happen.
In this way previous frameworks implicitly encoded \emph{negated} information.
Similarly, the a dataflow analysis producing an upper bound which increases throughout the analysis, such as a may-alias dataflow (\S~\ref{alias:sec:dataflow}), is not meaningful until it can no longer grow.
As a result, the usability of a dataflow analysis depends on the \emph{negated} information that no larger upper bound will be created, normally determined by waiting for the analysis to complete. 

Co-dependence eliminates the ability to stage analyses since it may not be possible to know that an analysis phase will not continue later.
As a result, the use of negated information must be made explicit (\S~\ref{holmes:sec:circ}).
Explicit negation also allows us to deal with more exotic control flow than ``to completion''
{\sc BiTR} performs simultaneous constraint solving for type inference, with a goal of ignoring as few constraints as possible while still arriving at a consistent answer.
Our form of negation is useful for describing that a particular solution is minimal (\S~\ref{bitr:sec:circ}), and avoiding the computation of non-minimal solutions.

\noindent\textbf{External Code.}
Datalog systems use external predicates to allow programmers to specify functionality outside the catalog programming language.
Access to external code is important to allow the use of already existing analysis code, such as BAP~\cite{bap} lifting and ELF parsing.
We allow a more limited form of external predicates we call \emph{callbacks}.
The use of callbacks as opposed to external predicates to more easily implement the engine (\S~\ref{holmes:sec:implcall}).
It also allows us to utilize traditional imperative data structures, such as hash tables, during the implementation of performance critical analyses (\S~\ref{chap:alias}).

We investigated our thesis statement by designing and implementing \sysname, a logic language engine designed for the integration of binary analyses.
BiTR (\S~\ref{chap:bitr}) inspired this design.
For this purpose we invented a novel form of negation, based on circumscription~\cite{circumscription} suited to this particular application (\S~\ref{holmes:sec:circ}).
This negation can further make progress from failed hypotheses (\S~\ref{holmes:sec:callcc}).
This adaptation allows Holmes to resolve an output even in the case of a negation cycle, as would be the case with a co-dependent control flow recovery and a value analysis assuming completeness of the control flow graph.
This is concretely the case in the use of VSA to resolve function pointers.
We also investigated different implementation approaches for \sysname (\S~\ref{holmes:sec:impl}).

We provide a formal semantics for the Holmes logic language (\S~\ref{chap:formal}).
These semantics enhance the ability to reason about the results of an analysis by formally constraining what derivations the analysis engine may follow.
They also make it possible to determine compliance for alternative implementations of the engine, potentially allowing programs to run on backends other than the initial one written for this thesis.

We show that \sysname can implement different sensitivities of alias analysis, and show how this affects practical use-after-free classification (\S~\ref{chap:alias}).
We implemented co-dependent alias analysis, use-after-free detection, and control flow recovery using \sysname.
This implementation provides evidence for the feasibility of using a logic language as an integration layer for cyclically dependent analyses.
This concrete bug finding tool demonstrates the applicability of the \sysname language to the domain of compiled-code program analysis.
We translate analysis techniques from the compiler and program analysis communities into analysis of compiled code (\S~\ref{alias:sec:system}).
We measure the performance cost and precision benefits of different sensitivities of alias analysis by leveraging the modularity of the \sysname-based implementation (\S~\ref{alias:sec:eval}).
We did not set out to invent a new or better alias analysis.
Rather, we set out to demonstrate the suitability of \sysname\ as a platform for this type of analysis.

We present a type recovery mechanism for compiled code which would be well suited to a Holmes-based implementation (\S~\ref{chap:bitr}).
This demonstrates that even work written without \sysname in mind may be naturally translatable into a Holmes implementation.
We define a \emph{descriptive} type system, more powerful than direct C types for performing inference, wile being more flexible than a traditional prescriptive type system (\S~\ref{sec:typesys}).
We address the problem of inconsistent type constraints by suggesting a resolution system which seeks a solution in which it drops the fewest constraints (\S~\ref{sec:infmeth}).
Finally, previous work in type recovery used a metric which minimized the importance of correct recovery of structure types.
In \S~\ref{subsec:metrics}, we provide a novel metric for type recovery based on the probability of correctness of any individual query by a reverse engineer or downstream analysis.
